---
title: "Lab: Choropleth Map"
author: "Xuemao Zhang"
date: "June 27, 2019"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    fontsize: 12pt
---

```{r setup, include=FALSE, warning=FALSE, error=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

## Introduction

* In this session, we use one example to show how to plot a choropleth map using the information for coloring all regions.

* We use the package `usmap` plot US map including Alaska and Hawaii. You may try to use the package `fiftystater`.

* Install the following packages if you don't have them.

```
                install.packages("devtools");
                devtools::install_github("wmurphyrd/fiftystater");
                install.packages("usmap");
                install.packages("dplyr");
                install.packages('ggplot2');
                install.packages('magick');
                install.packages('stringr');
``` 

* Consider the state and county unemployment rates data in US from 1990 to 2017.  The data are available from [Bureau of Labor Statistics](https://www.bls.gov/lau/).


## Unemployment Data

### State Data

- The state unemployment data (2013-2017) are in the form of HTML tables on https://www.bls.gov/lau/:

      - https://www.bls.gov/lau/lastrk13.htm
      - https://www.bls.gov/lau/lastrk14.htm
      - https://www.bls.gov/lau/lastrk15.htm
      - https://www.bls.gov/lau/lastrk16.htm
      - https://www.bls.gov/lau/lastrk17.htm

- The data were downloaded and combined in the file `StateUnemployment.csv`.

```{r echo=TRUE}
StateRate = read.csv("F:/DataCamp/data/StateUnemployment.csv",header=TRUE, sep=",");
dim(StateRate);
str(StateRate);
```

### County Data


- The county data from 1990 to 2017 were downloaded and combined in the file `CountyUnemployment.csv`.

```{r echo=TRUE}
library(dplyr);
CountyRate = read.csv("F:/DataCamp/data/CountyUnemployment.csv",header=TRUE, sep=",",stringsAsFactors = FALSE);
dim(CountyRate);
str(CountyRate);
CountyRate$Rate=as.numeric(CountyRate$Rate);
which(complete.cases(CountyRate)==F);  #some counties in some year have missing values
```


## Visualization of State Data

### Plot the data by year

- For example, we plot the unemployment rate data for the year 2017.

```{r fig.align="center", out.width = '100%', echo=TRUE}
library(usmap); 
library(ggplot2);
library(dplyr);
StateRate=dplyr::rename(StateRate, state=State); #rename "State" to "state" to use the usmap package
StateRate_2017=dplyr::filter(StateRate,Year==2017);
plot_usmap(data = StateRate_2017, values = "Rate", lines = "red") +
scale_fill_continuous(low = "green", high = "brown",
name = "Unemployment Rate (2017)", label = scales::comma)+
theme(legend.position = "top");
```

- Now,  we plot the unemployment rate data by Year. For this, we define a function of year.

```{r fig.align="center", out.width = '100%', echo=TRUE}
PlotRate= function(x)
{
data_year=dplyr::filter(StateRate, Year == x);
plot_usmap(data = data_year, values = "Rate", lines = "red") +
scale_fill_continuous(low = "green", high = "brown", name = "Rate", label = scales::comma)+
ggtitle(paste("Distribution of US Unemployment Rate in Year",x))+
theme(legend.position = "top");
}
```

- Show all 5 maps.

```{r fig.align="center", out.width = '100%', echo=TRUE}
for( i in 1:5)
{
print(PlotRate(i+2012));
}
```

### Animated maps

```{r fig.align="center", out.width = '100%', echo=TRUE}
library(magick);
img=image_graph(width=600,height=400,res=96);
for( i in 1:5) print(PlotRate(i+2012));
dev.off();
StateAnimation = image_animate(img, fps = 1);
print(StateAnimation);
```


## Visualization of County Data

- Add a variable "state".  You may follow the last section to plot the state level data.

```{r echo=TRUE}
library(stringr);  library(dplyr);
CountyRate=mutate(CountyRate, state=stringr::str_sub(County, -2, -1));  #extract the state name: the last two characters in "County"
head(CountyRate);
```

- Add the variable 'fips' for county fips following the last lecture.

```{r echo=TRUE}
library(dplyr);
CountyRate$StateFips=as.character(CountyRate$StateFips);
CountyRate$CountyFips=as.character(CountyRate$CountyFips);
CountyRate=mutate( CountyRate, fips=ifelse(nchar(CountyFips)==3,
paste(StateFips, CountyFips, sep=""),
ifelse(nchar(CountyFips)==2,
paste(StateFips, CountyFips, sep="0"),
paste(StateFips, CountyFips, sep="00")))  );
head(CountyRate);
``` 

- We first divide the unemployment rate into **seven** categories: 
    -  "<=2%"
    -  ">2% and <=4%"
    -  ">4% and <=6%"
    -  ">6% and <=8%"
    -  ">8% and <=10%"
    -  ">10% and <=12%"
    -  ">12%"

```{r echo=TRUE}
CountyRate= dplyr::mutate(CountyRate,
rate = ifelse(Rate<=2.0, "2% or lower", 
       ifelse(Rate>2 & Rate<=4, "2% to 4%",
       ifelse(Rate>4 & Rate<=6, "4% to 6%",
       ifelse(Rate>6 & Rate<=8, "6% to 8%",
       ifelse(Rate>8 & Rate<=10, "8% to 10%",
       ifelse(Rate>10 & Rate<=12, "10% to 12%", "12% or over"
       ))))))
    );
CountyRate$rate=as.factor(CountyRate$rate);
CountyRate$rate=factor(CountyRate$rate, levels = c("12% or over","10% to 12%","8% to 10%","6% to 8%","4% to 6%","2% to 4%", "2% or lower")); #order the factor levels
which(complete.cases(CountyRate$rate)==F);
```


- Now we plot the county data by Year. For this, we define a function of year.


```{r echo=TRUE}
library(usmap);  library(ggplot2);
PlotCountyRate= function(x)
{
County_year=dplyr::filter(CountyRate, Year==x);
plot_usmap(regions="counties",data = County_year, values ="rate", lines=NA) +
ggplot2::scale_fill_brewer(palette="Accent", name ="Unemployment Rate", na.value = "black")+
ggtitle(paste("Distribution of US Unemployment Rate by County in Year",x))+
theme(legend.position = "right");
}
```

- Animated maps.

```{r fig.align="center", out.width = '80%', echo=TRUE}
library(magick);
img=image_graph(width=600,height=400,res=96);
for( i in 1:28) print(PlotCountyRate(i+1989));
dev.off();
CountyRateAnimation = image_animate(img, fps = 1);
print(CountyRateAnimation);
image_write(CountyRateAnimation, "CountyRateAnimation.gif")
```